Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-07-29 analysis/paper/paper.Rmd
Cutting down text. Still need table of chronological phases

```{r wrangling}
sites <- assemblages %>% 
      # Find mean reported date for each site (to be used in the plots below)
  rowwise() %>% 
  mutate(mean_date = mean(c(reported_start_bc, reported_end_bc))) %>% 
  mutate(
    # Find total artefact count and flint count for each site
    artefact_count = sum(c_across(core:flake_retouch_other)),
    flint_count = sum(c_across(core:flake_retouch)),
    nonflint_count = sum(c_across(axe_other:flake_retouch_other))) %>%
  # Exclude sites with less than 100 artefacts
  filter(artefact_count > 100) %>% 
  arrange(mean_date) %>% 
  rownames_to_column(var = "no")

@@

The Early Mesolithic, or Flake Axe Phase, is defined as lasting from c. 9200--8200 BCE [@reitan2016] and is set to start with the first recorded human presence in Norway. Previous research has typically proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g. @bjerck2008; @breivik2016; @fuglestvedt2012; but see @viken2018c]. Around the transition to the subsequent Middle Mesolithic, or Microlith Phase, at 8200 BCE pervasive changes in blade and axe technology takaes place [@damlien2016; @eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian steppes [@günther2018; @manninen2021]. The Microlith Phase is defined as lasting until around 7000 BCE, which is followed by the Pecked Adze Phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016]. The next typological transition at c. 5600 BCE signifies the onset of the Nøstvet Adze Phase. While previously defined as having a slightly longer duration, the Nøstvet Phase has traditionally been seen as representing the onset of more varied settlement systems and stable mobility patterns [e.g. @jaksland2001; @lindblom1984]. In recent years it has been suggested that the transition to a decrease in mobility and more varied land-use patterns can be traced back to the Middle Mesolithic [@solheim2016]. The subsequent Transverse Arrowhead Phase (c. 4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of new flint projectiles [@reitan2016]. It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c. 4300 BCE [@nielsen2021].

@@

# Combine spatial point data with assemblage data (holds excavation project
# for colouring points in the maps). Set up aggregate labels. There might be
# a way to avoid having to do this manually, for example by using coordinates?
sitejoin <- right_join(site_pts, sites, by = c("LOKID" = "askeladden_id")) %>% 
  mutate(no = as.numeric(no)) %>% 
    mutate(no_m = ifelse(no %in% c(1:10) & !duplicated(no %in% c(1:10)),
                         "1-10",
           ifelse(no %in% 11:12 & !duplicated(no %in% 11:12), "11-12", 
           ifelse(no %in% 14:17 & !duplicated(no %in% 14:17), "14-17", 
           ifelse(no %in% c(21:22, 44) & !duplicated(no %in% c(21:22, 44)),
                  "21-22, 44", 
           ifelse(no %in% c(25:27, 30) & !duplicated(no %in% c(25:27, 30)),
                  "25-27, 30",
           ifelse(no %in% c(36, 41) &
                    !duplicated(no %in% c(36, 41)),
                  "36, 41",        
           ifelse(no %in% c(19, 28:29, 23) & 
                !duplicated(no %in% c(19, 28:29, 23)),
                "19, 23, 28-29",
           ifelse(no %in% c(32:33, 35) & 
                !duplicated(no %in% c(32:33, 35)),
                "32-33, 35",
           ifelse(no %in% c(37:40)  &
                    !duplicated(no %in% c(37:40)), 
                  "37-40",
           ifelse(no %in% c(24, 34)  &
                    !duplicated(no %in% c(24, 34)), 
                  "24, 34",
           ifelse(no %in% c(42:43)  &
                    !duplicated(no %in% c(42:43)), 
                  "42-43",
          ifelse(no %in% c(45:46, 53)  &
                    !duplicated(no %in% c(45:46, 53)), 
                  "45-46, 53",
           ifelse(no %in% c(48:50, 54) & !duplicated(no %in% c(48:50, 54)),
                  "48-50, 54",
           ifelse(no %in% c(31, 47, 51:52) &
                    !duplicated(no %in% c(31, 47, 51:52)),
                  "31, 47, 51-52",
           ifelse(no %in% c(13, 20), no, ""))))))))))))))))

# Assemble site map. anc defined first is used for placing the scalebar,
# which required some tweaking to get right.
anc <- as.numeric(c(sitbbox2$ymin, sitbbox2$xmax))
sitemap <-
  ggplot() +
  geom_sf(data = count_reproj2, fill = "grey", colour = NA) +
  geom_sf(data = sitejoin, aes(fill = excavation_project, label = no),
          size = 2.25, shape = 21, 
          colour = "black") +
  ggrepel::geom_text_repel(data = sitejoin, aes(label = no_m,
                                                 geometry = geom),
                                                 stat = "sf_coordinates",
                           box.padding = 0.2, 
                           max.overlaps = getOption("ggrepel.max.overlaps",
                                                    default = 54)) +
  labs(fill = "Excavation project") +
  scale_fill_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
                    values = c("springgreen4", "royalblue2",
                                "gold1", "firebrick3")) +
  ggsn::scalebar(data = site_pts, dist = 10, dist_unit = "km", 
                 transform = FALSE, st.size = 4, height = 0.05, 
                 border.size = 0.05, st.dist = 0.07,
                 anchor = c(x = anc[2] - 8000, y = anc[1]) + 4000) +
  coord_sf(xlim = c(sitbbox2[1], sitbbox2[3]), 
           ylim = c(sitbbox2[2], sitbbox2[4]),
         expand = FALSE) +
 theme_bw() + theme(axis.title=element_blank(),
                     axis.text.y = element_blank(),
                     axis.text.x = element_blank(),
                     rect = element_rect(),
                     axis.ticks = element_blank(),
                     panel.grid.major = element_blank(),
                     legend.position = "none") +
  guides(fill = guide_legend(ncol = 2,
                             title.position = "top", title.hjust = 0.5))

@@

sites %>% 
  mutate(strt_bce = reported_start_bc * -1, 
         end_bce = reported_end_bc * -1,
         dating_method = ifelse(dating_method == "shore_typo", "Shoreline/typology", "Radiocarbon")) %>% 
  select(no, site, dating_method, strt_bce, end_bce) %>% 
  gt() %>% 
    cols_label(
    site = "Site name",
    dating_method = "Dating method",
    strt_bce = "Reported start (BCE)",
    end_bce = "Reported end (BCE)",
  ) %>% 
  tab_header(
    title = "Table 1. Analysed sites."
  ) %>% 
  tab_options(heading.title.font.size = 15)
```

@@

The general impression from the CA is that a chronological dimension is associated with the patterning in the data (Figure \@ref(fig:cor)). This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data, as well as the horseshoe curve or Guttman effect evident in the column plot [Figure \@ref(fig:cor)B, @baxter1994, 119--120]. The earliest sites tend to be located in the upper left corner of plot A, with increasingly younger sites towards the bottom along the second dimension. The sites from the later parts of the Mesolithic are drawn out along the first dimension of the plot, and are not as impacted by the second dimension as the more numerous older sites.

The column plot (Figure \@ref(fig:cor)B) reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage. It is also interesting that these sites to a larger extent are characterised by core fragments, both in flint and non-flint materials, rather than the cores themselves. The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites.

The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages. Sites with high values on the first dimensions are later Mesolithic sites associated with axe production in non-flint materials, but the later sites occur along the entire dimension, indicating that while these axe production sites are a feature of the later Mesolithic, there is marked variation among the sites. Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious early case with an assemblage characterised by axe production in metarhyolite [@eigeland2014]. However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014 [124] to suggest that the material might have been compromised. This could explain its position as an outlier in the plot. Finally, although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.

@@

Thus, while some secondary expectations of the WABI does not seem to apply to the present material, it is difficult to say to what degree this is caused by idiosyncrasies in the Norwegian system for classification of lithics and properties of the lithic material itself. The relationship between VDL and RFSL does correspond to the model and follows a clear temporal trend that is also correlated with the increased use of local raw material. Thus, if the relationship between VDL and RFSL is accepted as a proxy for curation, and is related to land-use and mobility patterns, these findings would be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases [e.g. @bergsvik2001; @bjerck2008; @glørstad2010; @jaksland2001]. To explore this proposition further, these two variables are subjected to more detailed scrutiny below.

@@

```{r wabi, fig.cap = "Relative frequency of secondarily worked lithics plotted against the volumetric density of artefacts (artefact count / excavated m^3^) for A) All lithics, B) Flint. The logarithm is taken to base 10 on all axes.", fig.height = 5, fig.width = 10}

@@

The results of the CA appear to align well with previous research [e.g. @solheim2017c, with references]. In the flint material the earliest sites are separated from the rest primarily based on the presence of macro tools, microburins, projectiles, and, for slightly younger sites, core fragments and blades [cf. @bjerck2017; @breivik2018; @fuglestvedt2007; @jaksland2014b; @damlien2018b]. The importance of the latter two can be associated with the blade technology that is introduced with the Middle Mesolithic, characterised by blade production from conical and sub-conical cores with faceted platforms that involves the removal of core tablets and rejuvenation flakes [@damlien2016]. When it comes to the non-flint material, projectiles are to a larger extent a property of the earlier sites than later ones. The use of metarhyolite for the production of axes is present at some earlier sites in addition to the previously mentioned Nedre Hobekk 2, and the production of hatches and core axes is introduced in the Microlith Phase [@eymundsson2018; @jaksland2014b; @reitan2016]. However, in agreement with the literature, this is evidently not as prominent a part of these assemblages.

The flint material of the later sites are to larger extent characterised by micro-blades, which corresponds to the transition to micro-blade production from handle cores [e.g. @solheim2020]. A more fragmented flint material, as indicated by the relative importance of flint chips and fragments, is also a previously noted property of some later Mesolithic, as well as early Neolithic sites [e.g. @fossum2017; @stokke2018a]. The most defining material for the later sites, however, is non-flint macro tools and associated debitage, which is dominating some of these assemblages. It was noted above that this material does not seem to impact the latest sites, which could indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous suggestions [e.g. @glørstad2011; @reitan2016].

@@

While there are certainly nuances in the material that might lead one to question the applicability of the VDL and RFSL measures for any individual site, the overall pattern for curation does appear relatively robust. The curation index is relatively high and uniform until some time before 8000 BCE. This corresponds well with the view that the Early Mesolithic is characterised by a high and uniform degree of mobility [e.g. @bjerck2008; @breivik2016; @fuglestvedt2012]. This is followed by a marked increase in expedience, which has stabilised by around 7000 BCE. Again, this corresponds well with the employed chronological framework. Referring back to the demographic changes that are to take place around this transition, the Microlith phase could thus represent a period where migrating people and new living practices were propagating through societies in south-eastern Norway---a process that in light of the curation data would have concluded around 7000 BCE.

@@

As it stands, the main hypotheses resulting from the present analysis would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high and uniform degrees of mobility, which then drop before levelling off at around 7000 BCE. These then remain relatively stable throughout the rest of the period, despite variation pertaining to other aspects of the lithic inventories, as evidenced by the CA. Although the precise nature of this transition would require further consideration, the quite dramatic fall in curation levels and parallel increase in variation would seem to correlate well with a transition from a predominantly residential to logistical settlement system.

@@

### Colophon

This report was generated on `r Sys.time()` using the following computational environment and dependencies:

```{r colophon, cache = FALSE}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

The current Git commit details are:

```{r}
# what commit is this file at? 
if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())  
```

>>>>

```{r wrangling}
sites <- assemblages %>% 
      # Find mean reported date for each site (to be used in the plots below)
  rowwise() %>% 
  mutate(mean_date = mean(c(reported_start_bc, reported_end_bc))) %>% 
  mutate(
    # Find total artefact count and flint count for each site
    artefact_count = sum(c_across(core:flake_retouch_other)),
    flint_count = sum(c_across(core:flake_retouch)),
    nonflint_count = sum(c_across(axe_other:flake_retouch_other))) %>%
  # Exclude sites with less than 100 artefacts
  filter(artefact_count > 100) %>% 
  arrange(mean_date) %>% 
  tibble::rownames_to_column(var = "no")

@@

The Early Mesolithic, or Flake Axe Phase, is defined as lasting from c. 9200--8200 BCE [@reitan2016] and is set to start with the first recorded human presence in Norway. Previous research has typically proposed that the Early Mesolithic is characterised by a relatively high degree of mobility, and low variation in site types and associated mobility patterns [e.g. @bjerck2008; @breivik2016; @fuglestvedt2012; but see @viken2018c]. Around the transition to the subsequent Middle Mesolithic or Microlith Phase at c. 8200 BCE, pervasive changes in blade and axe technology occur [@damlien2016; @eymundsson2018; @solheim2020], which in turn has been associated with changes in population genomics and related migration events hailing from the Eurasian steppes [@günther2018; @manninen2021]. The Microlith Phase is defined as lasting until around 7000 BCE, which is followed by the Pecked Adze Phase, characterised by a more dominating presence of non-flint macro tools and associated production waste in the assemblages [@reitan2016]. The next typological transition at c. 5600 BCE signifies the onset of the Nøstvet Adze Phase. While previously defined as having a slightly longer duration, the Nøstvet Phase has traditionally been seen as representing the onset of more varied settlement systems and stable mobility patterns [e.g. @jaksland2001; @lindblom1984]. In recent years it has been suggested that the transition to a decrease in mobility and more varied land-use patterns can be traced back to the Middle Mesolithic [@solheim2016]. The subsequent Transverse Arrowhead Phase (c. 4500--3900 BCE) is characterised by a dramatic decrease in axe finds, and the introduction of new flint projectiles [@reitan2016]. It has recently been suggested that a dispersal of people from southern Scandinavia into southern Norway takes place in this period [@eigeland2015, 379; @nielsen2021], which could follow after a preceding population decline at c. 4300 BCE [@nielsen2021].

@@

# Combine spatial point data with assemblage data (holds excavation project
# for colouring points in the maps). Set up aggregate labels. There might be
# a way to avoid having to do this manually, for example by using coordinates?
sitejoin <- right_join(site_pts, sites, by = c("LOKID" = "askeladden_id"))

# Assemble site map. anc defined first is used for placing the scalebar,
# which required some tweaking to get right.
anc <- as.numeric(c(sitbbox2$ymin, sitbbox2$xmax))
sitemap <-
  ggplot() +
  geom_sf(data = count_reproj2, fill = "grey", colour = NA) +
  geom_sf(data = sitejoin, aes(fill = excavation_project),
          size = 2.25, shape = 21, 
          colour = "black") +
  labs(fill = "Excavation project") +
  scale_fill_manual(labels = c("E18 Bommestad-Sky", "E18 Brunlanes",
                               "E18 Rugtvedt-Dørdal", "Vestfoldbaneprosjektet"),
                    values = c("springgreen4", "royalblue2",
                                "gold1", "firebrick3")) +
  ggsn::scalebar(data = site_pts, dist = 10, dist_unit = "km", 
                 transform = FALSE, st.size = 4, height = 0.05, 
                 border.size = 0.05, st.dist = 0.07,
                 anchor = c(x = anc[2] - 8000, y = anc[1]) + 4000) +
  coord_sf(xlim = c(sitbbox2[1], sitbbox2[3]), 
           ylim = c(sitbbox2[2], sitbbox2[4]),
         expand = FALSE) +
 theme_bw() + theme(axis.title=element_blank(),
                     axis.text.y = element_blank(),
                     axis.text.x = element_blank(),
                     rect = element_rect(),
                     axis.ticks = element_blank(),
                     panel.grid.major = element_blank(),
                     legend.position = "none") +
  guides(fill = guide_legend(ncol = 2,
                             title.position = "top", title.hjust = 0.5))

@@

sites %>% 
  mutate(strt_bce = reported_start_bc * -1, 
         end_bce = reported_end_bc * -1,
         dating_method = ifelse(dating_method == "shore_typo", "Shoreline/typology", "Radiocarbon")) %>% 
  select(no, site, dating_method, strt_bce, end_bce) %>% 
  gt() %>% 
    cols_label(
    site = "Site name",
    dating_method = "Dating method",
    strt_bce = "Reported start (BCE)",
    end_bce = "Reported end (BCE)",
  ) %>% 
  tab_header(
    title = "Table 1. Analysed sites."
  ) %>% 
  tab_options(heading.title.font.size = 13, heading.align = "left")
```

@@

The general impression from the CA is that a chronological dimension is associated with the patterning in the data (Figure \@ref(fig:cor)). This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), the fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)` % of the inertia or variance in the data, as well as the horseshoe curve or Guttman effect evident in the column plot [Figure \@ref(fig:cor)B, @baxter1994, 119--120].

The column plot reveals that the earliest sites are characterised by the flint artefact categories microburins, projectiles, as well as flint macro tools and associated debitage. It is also interesting that these sites to a larger extent are characterised by core fragments, both in flint and non-flint materials, rather than the cores themselves. The non-flint material on the earliest, or among the earliest sites, appears to be centred around the production of projectiles, as both the projectiles themselves and non-flint blades are important constituents of the assemblages at these sites. The first dimension, which is pulling some of the later sites towards the right of the plot, is mainly defined by macro tools and associated debitage in non-flint materials that are negatively correlated with more flint dominated assemblages. Site number 9, Nedre Hobekk 2, located in the upper right quadrant of the row plot represents a somewhat curious case in that it is an early assemblage characterised by axe production in metarhyolite [@eigeland2014]. However, as the site had been quite heavily impacted by modern disturbances, this led @eigeland2014 [124] to suggest that the material might have been compromised. This could explain its position as an outlier in the plot. Finally, although the sample size is quite strained and the discussion of finer chronological points might not be warranted, the first dimension does appear to be of of less importance for the absolute latest sites, as indicated by their location to the left of the plot.

@@

Thus, while some secondary expectations of the WABI do not seem to apply to the present material, it is difficult to say to what degree this is caused by idiosyncrasies in the Norwegian system for classification of lithics and properties of the lithic material itself. The relationship between VDL and RFSL does correspond to the model and follows a clear temporal trend that is also correlated with the increased use of local raw material. Thus, if the relationship between VDL and RFSL is accepted as a proxy for curation, and is related to land-use and mobility patterns, these findings would be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases [e.g. @bergsvik2001; @bjerck2008; @glørstad2010; @jaksland2001]. To explore this proposition further, these two variables are subjected to more detailed scrutiny below.

@@

```{r wabi, fig.cap = "Relative frequency of secondarily worked lithics plotted against the volumetric density of artefacts for A) All lithics, B) Flint. The logarithm is taken to base 10 on all axes.", fig.height = 5, fig.width = 10}

@@

The results of the CA appear to align well with previous research [e.g. @solheim2017c, with references]. In the flint material the earliest sites are separated from the rest primarily based on the presence of macro tools, microburins, projectiles, and, for slightly younger sites, core fragments and blades [cf. @bjerck2017; @breivik2018; @fuglestvedt2007; @jaksland2014b; @damlien2018b]. The importance of the latter two can be associated with the blade technology that is introduced with the Middle Mesolithic, characterised by blade production from conical and sub-conical cores with faceted platforms that involves the removal of core tablets and rejuvenation flakes [@damlien2016]. When it comes to the non-flint material, projectiles are to a larger extent a property of the earlier sites than later ones. The use of metarhyolite for the production of axes is present at some earlier sites in addition to the previously mentioned Nedre Hobekk 2, and the production of non-flint hatches and core axes is introduced in the Microlith Phase [@eymundsson2018; @jaksland2014b; @reitan2016]. However, in agreement with the literature, this is evidently not as prominent a part of these assemblages.

The flint material of the later sites are to larger extent characterised by micro-blades, which corresponds to the transition to micro-blade production from handle cores [e.g. @solheim2020]. A more fragmented flint material, as indicated by the relative importance of flint chips and fragments, is also a previously noted property of some later Mesolithic, as well as early Neolithic sites [e.g. @fossum2017; @stokke2018a]. The most defining material for the later sites, however, is non-flint macro tools and associated debitage, which is dominating some of these assemblages. It was noted above that this material does not seem to impact the latest sites, which would indicate that specialised axe production sites disappear towards the end of the Mesolithic, a notion that would be in line with previous suggestions [e.g. @glørstad2011; @reitan2016].

@@

While there are certainly nuances in the material that might lead one to question the applicability of the VDL and RFSL measures for any individual site, the overall pattern for curation does appear relatively robust. The curation index is relatively high and uniform until some time before 8000 BCE. This corresponds well with the view that the Early Mesolithic is characterised by a high and uniform degree of mobility. This is followed by a marked increase in expedience, which has stabilised by around 7000 BCE. Again, this corresponds well with the employed chronological framework. Referring back to the demographic changes that are to take place around this transition, the Microlith phase could thus represent a period where migrating people and new living practices were propagating through societies in south-eastern Norway---a process that in light of the curation data would have concluded around 7000 BCE.

@@

As it stands, the main hypotheses resulting from the present analysis would be that settlement patterns in the earliest parts of the Mesolithic were characterised by relatively high and uniform degrees of mobility, which then drop before levelling off at around 7000 BCE. These then remain relatively stable throughout the rest of the period, despite variation pertaining to other aspects of the lithic inventories, as evidenced by the CA. Although the precise nature of this transition would require further consideration, the fall in curation levels and parallel increase in variation would seem to correlate well with a transition from a predominantly residential to logistical settlement system.

@@

<!-- \newpage -->

<!-- ### Colophon -->

<!-- This report was generated on `r Sys.time()` using the following computational environment and dependencies: -->

<!-- ```{r colophon, cache = FALSE} -->
<!-- # which R packages and versions? -->
<!-- if ("devtools" %in% installed.packages()) devtools::session_info() -->
<!-- ``` -->

<!-- The current Git commit details are: -->

<!-- ```{r} -->
<!-- # what commit is this file at?  -->
<!-- if ("git2r" %in% installed.packages() & git2r::in_repository(path = ".")) git2r::repository(here::here())   -->
<!-- ``` -->
